---
title: "HW6"
output: 
 html_notebook:
    toc: true
    toc_float: true
    number_sections: true
---
Use this R Notebook document to answer the questions and document your work. 
Enter the R code used to answer each question in the corresponding R code chunk. 
Write any textual explanations **outside** of the chunks. When we grade, we will *hide* the code and just look at the R output and your written answers.
Attempt to clean up your code as much as possible so that only the necessary lines remain. 

When you are done:

1. Select 'Run All' from the 'Run' dropdown menue.
2. Save (File -> Save)
3. Click 'Preview' to bring up the `HW1.nb.html` file. Check through this to make sure it rendered correctly.
4. Upload the `HW1.nb.html` to Canvas.

Re-install PLS205 package: 
Run the following command directly in your RStudio console:
`devtools::install_github('deruncie/PLS205_package')`

Necessary libraries
```{r echo = F,message=FALSE, warning=FALSE} 
# Necessary packages 
library(ggplot2) 
library(emmeans)
library(lme4)
library(lmerTest)
library(PLS205)
```

## Question 1

A researcher was interested in examining the effect of pasture species composition on beta-carotene content in eggs of pastured poultry. 
She evenly divided each of four fields into five sections and overseeded each section with one of the following: 
1) annual rye (grass); 2) fescue (grass); 3) white clover (legume); 4) vetch (legume); or, 5) clover/fescue mix. 
A separate flock of 10 birds was put out in each section. 
After one-week, eggs from each flock were collected daily and combined for determining beta-carotene content of the yolks. 
These values in $\mu g/g$ were averaged after 10 days.

```{r echo=FALSE}
data_1 = read.csv('egg_carotene.csv')
str(data_1)
```

### 1.1 Describe the design of this experiment in detail.

**Design**: Complete Randomized(?) Block Design
**Response**: beta-carotene content

| Structure | Variable      | Type        | # levels | EU            
|-----------|---------------|-------------|----------|---------------
| Treatment | Species       | Categorical | 5        | Field:Species 
| Design    | Field         | Categorical | 4        |               
|           | Field:Species | Categorical | 20       |               
| Response  | Beta          | Numeric     | 20       |               

> Each field is a block, with the cover species treatments applied to five separate sections within a block. I assume they are assigned randomly. The chickens are not included in the data, possibly because they are aliased with Field:Section and Field:Species.


### 1.2 Assess whether the assumptions of a linear model are adequately satisfied. Is a transformation needed?
    
```{r}
# setting up the data
data_1$Field <- as.factor(data_1$Field)
data_1$Section <- as.factor(data_1$Section)
data_1$Species <- as.factor(data_1$Species)

# checking aliases, crosses, nests
# NOTE: Can't get is_aliased() to work after updating the PLS package...
is_aliased(Field ~ Species, data = data_1)

is_crossed(Beta ~ Field, data = data_1)

is_nested
# creating the model
egg_model <- lm(Beta ~ Species + Field, data = data_1)

# linearity and normality diagnostics
pls205_diagnostics(egg_model, EU = 'Field:Species')

```

> The EU plot seems to be reasonably linear. The points begin to trail off at the top and there is a "dip" in the middle, so a transformation may achieve a better fit. However, I believe that a transformation would only marginally improve our analyis.
> The Scale-Location plot has a decent spread, but there are more larger in the middle values. A tranformation might improve this fit, but again, the improvement would likely be marginal.

### 1.3 Estimate the differences between all pairs of pasture covers. Include appropriate units based on your choice of transformation (or no transformation)

```{r}

```

> ENTER YOUR RESPONSE HERE most not significant, emmeans contrast methods pairwise

### 1.4 Is there any evidence of an interaction between Field and Species cover? ie. do differences among covers change across the fields?
Is the interaction replicated? If not, you can't answer this question with an ANOVA, but you can tentatively assess with an interaction plot. 
```{r}
```

> ENTER YOUR RESPONSE HERE


### 1.5 Did the researcher gain information by utilizing a RCBD relative to a CRD where the sections of all fields were randomized together?
```{r}
```

> ENTER YOUR RESPONSE HERE

### 1.6 Instead of randomizing all 5 plots on the 4 fields together, speculate on the efficiency had she used only one field.
In this case, to maintain the same replication, she would divide it into 20 smaller plots, and use only 2 birds per plot (instead of 10).
Speculate on how the MSE would compare to the original RCBD, or the CRD over the four fields. 
**Is there any advantage to spreading the experiment over 4 fields** Is there any disadvantage?

> ENTER YOUR RESPONSE HERE

### 1.7 Add 1 $\mu g/g$ to each value in Field F4. Does this produce any effect on the SS_Pasture, MS_Pasture, or F-value? Explain why or why not in one sentence.
In this case (with Field F4 being much more productive), did the importance of Blocking increase or decrease?

The following code makes a modifed dataset with the increase of 1 in the Beta values for Field F4

```{r}
data_2_modified = data_1
data_2_modified$Beta[data_2_modified$Field == 'F4'] = data_2_modified$Beta[data_2_modified$Field == 'F4'] + 1
```

> ENTER YOUR RESPONSE HERE
