---
title: "HW2"
output:
  html_notebook:
    number_sections: yes
    toc: yes
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
---

Use this R Notebook document to answer the questions and document your work. 
Enter the R code used to answer each question in the corresponding R code chunk. 
Write any textual explanations **outside** of the chunks. When we grade, we will *hide* the code and just look at the R output and your written answers.
Attempt to clean up your code as much as possible so that only the necessary lines remain. 

When you are done:

1. Select 'Run All' from the 'Run' dropdown menue.
2. Save (File -> Save)
3. Click 'Preview' to bring up the `HW2.nb.html` file. Check through this to make sure it rendered correctly.
4. Upload the files: `HW2.nb.html` and `HW2.Rmd` to Canvas.

---

# Experiment
A researcher is interested in the difference in root growth between two wheat varieties. She plants
16 seeds of each variety in a randomized design and harvests the roots after 1 week. She records
the root length (cm) of each plant.

```{r echo = FALSE}
# This loads the data for this question
data_1 <- read.csv('~/Desktop/R_Projects/PLS205_Winter2021/PLS205_Winter2021/data/wheat_roots.csv')
# This prints a summary of the data table
str(data_1)

data_1$Plant <- as.character(data_1$Plant)
summary(data_1)

```

## Describe the design of this experiment in detail. 

Use the following table. When you click `Preview`, RStudio will format this table into a nice HTML table.
You can also fill out the table in an excel document, and then paste it into [this website](http://www.tablesgenerator.com/markdown_tables) to generate the Markdown table. 

**Design**: Completely randomized design

| Structure | Variable    | Type        | # levels | Experimental Unit |
|-----------|-------------|-------------|----------|-------------------|
| Treatment | Variety     | Character   | 2        |  Plant            |
| Design    | Plant       | Numeric     | 32       |                   |
| Response  | Root Length | Numeric     | 32       |                   |


## To help you analyze this data, I've separated out the data from each variety into a vector:
```{r}
Variety_A <- data_1$Root_length[data_1$Variety == 'A']
Variety_B <- data_1$Root_length[data_1$Variety == 'B']
```

Use these variables below for your calculations.

## Estimate the difference in root lengths between the two varieties
You can check your answers below with the `t.test` function, but you must write out the calculations by hand. You can use functions like `mean()` and `sd()`, `pnorm` --> `pt`, `qt`, etc.
```{r}
data_1_diff <- mean(Variety_B) - mean(Variety_A)
data_1_diff

```

> The difference in root lengths is 3.47 cm.

## Estimate the standard error of this difference. Assume the variances of the two varieties are the same, and find a pooled estimate of the variance.
```{r}
# Variance

# Using var() to determine the variance for Varieties A and B
data_1_var_A <- var(Variety_A)
data_1_var_A
data_1_var_B <- var(Variety_B)
data_1_var_B

# Pooling the variance
data_1_pooled_var <- ((16-1) * data_1_var_A + (16-1) * data_1_var_B) / ((16-1) + (16-1))
data_1_pooled_var

# Calculating SED
data_1_SED <- sqrt( sd(Variety_A)^2 / 15 + sd(Variety_B)^2 / 15)
data_1_SED

```

> data_1_SED = 0.84 cm

> data_1_pooled_var = 5.34 cm^2

## Is it likely to have observed this large a difference by chance if there were actually no difference?
```{r}

SE <- data_1_SED / sqrt(16)
SE

alpha <- 0.05
t_crit = qt(p = 1-alpha/2, df = 16-1)
t_crit

CI <- data_1_diff + c(-t_crit * SE, t_crit * SE)
CI


# ^ Come back to this question!!!!!!!!!!! **********
```

> *************ENTER YOUR RESPONSE HERE

## Repeat the analysis using the `lm` and `emmeans` functions. Do you get the same answer?
```{r}
library(emmeans)

lm(data_1$Root_length ~ data_1$Variety, data = data_1)

means_model1 <- emmeans(model_1,spec = 'Variety') # prep the emmeans
summary(means_model1,level = 0.95, infer = c(T,F)) # show confidence intervals

```

> ENTER YOUR RESPONSE HERE

## Make the model diagnostic plots shown in lab. Is there reason to be concerned about any of the model assumptions?
```{r}


```

> ENTER YOUR RESPONSE HERE

## Calculate the power of the test to detect a difference of 3cm, with n=16 and alpha = 0.1. Assume the within-variety standard deviation is is 2.5cm
```{r}
# see: ?power.t.test

```

> ENTER YOUR RESPONSE HERE

## Prepare a table showing the number of replicates required to detect significant differences (alpha = 0.01, power = 0.9) between means that are 1, 2, 4, 8, or 16 cm apart, assuming that the population standard deviation is 4cm. Use the function `power.t.test`. How much easier is it to detect a difference of 16cm than one of 1 cm?
```{r}
delta = c(1,2,4,8,16)
reps = c(NA,NA,NA,NA,NA)

# Here's a hit at how to calculate the sample size for the first delta
# reps[1] <- power.t.test(n = ?,  # make sure to fill in ALL of the ?
#                         delta = delta[1],
#                         sd = ?,
#                         sig.level = ?,
#                         power = ?,
#                         type = ?,
#                         alternative = ?
#                           )$n

# Here we make a table of the results. 
ans_4.4 <- data.frame(delta,reps)
ans_4.4
```
> ENTER YOUR RESPONSE HERE

