---
title: "HW5"
output:
  html_notebook:
    toc: yes
    toc_float: yes
    number_sections: no
  html_document:
    toc: yes
    df_print: paged
---

> Grading Notes:
>
> For questions that require a text answer:
> Full credit: requires an answer to each question asked. Your explanations need to valid
and logical. You don't need to use the same words as the key, but the meaning needs to be complete.
> Assign partial credit for answers that are incomplete. For multi-part questions, divide points
approximatly equally unless otherwise specified.
>
> For all questions, if you get the right answer using a different approach than I use, that is fine.

> Fill in your scores in the google form here: https://forms.gle/oxWPew8aS9Ds6bW49. 
PLEASE ONLY ENTER A NUMBER. IF YOU WANT TO LEAVE A COMMENT ABOUT A SCORE, DO IT DIRECTLY ON CANVES.
It will require you to log in with your ucdavis email address. Let me know if this causes problems. This way I can associate your grades with you on Canvas


Install new PLS205 package: 
```{r}
# if you get an error with pls205_diagnostics, re-install the package here
# devtools::install_github('deruncie/PLS205_package')
```

```{r}
library(ggplot2)
library(emmeans)
library(PLS205)
library(lme4)
library(lmerTest)
```


This homework follows from Lab5 in investigating the effect of Nitrogen on potato yield.

The code below repeats the trend analysis of this experiment presented in lab. 
We'll call this experimental design **Experiment 1**.

## Experiment 1

An experiment was run to evaluate effects of increased nitrogen fertilization on tuber yield of Alpine Russet potatoes

- 5 nitrogen regimes: 0, 90, 150, 210, 270 lbs / acre at emergence
- 10 reps (plots) / Nitrogen level
- Response: total yield
  
**Design Table**

| Structure | Variable  | Type        | # levels | EU        |
|-----------|-----------|-------------|----------|-----------|
| Treatment | Nitrogen  | Numeric     | 5        | NitrogenF |
| Design    | NitrogenF | Categorical | 5        |           |
|           | Plot      | Categorical | 50       |           |
| Response  | Yield     | Numeric     | 50       |           |

Load the data and define a categorical variable for the EU:
```{r}
experiment1 = read.csv('~/Desktop/R_Projects/PLS205_Winter2021/PLS205_Winter2021/data/Experiment1.csv')
experiment1$NitrogenF = as.factor(experiment1$Nitrogen)
str(experiment1)
```

Fit a degree-2 trend model
```{r}
expt1_deg2 = lmer(Yield ~ I(Nitrogen) + I(Nitrogen^2) + (1|NitrogenF), data = experiment1)
```

Model diagnostics:
```{r}
pls205_diagnostics(expt1_deg2,EU = 'NitrogenF')
```

Visualize the fit and the confidence interval:
```{r}
# calculate the trend at a range of values of Nitrogen:
Nitrogen_range = seq(0,320,by=1)
expt1_deg2_means = emmeans(expt1_deg2,spec = 'Nitrogen',at=list(Nitrogen = Nitrogen_range))
expt1_deg2_means_summary = as.data.frame(summary(expt1_deg2_means,infer = c(T,F),level = 0.95,ddf='K'))
# note: I forgot to add the ddf='K' in the lab. I've included it here for reference, but don't take off if you don't have it.

# Plot the trend and its confidence intervals:
#   Inputs: 
#      - `experiment1` the data table of the raw data
#      - `expt1_deg2_means_summary` table of estimates (emmean) and confidence intervals at each value of Nitrogen_range
#      * be sure to change these variable names in the code below to plot trends for other data/models

# Make a base plot with boxplots of the data
base_plot = ggplot(experiment1,aes(x=Nitrogen)) + 
    geom_boxplot(aes(y=Yield,group = Nitrogen),color = 'blue',position = position_identity()) + 
    ggtitle('Alpine Russet') + expand_limits(y=0)

# Then, add a line for the trend and a ribbon for the confidence intervals.
# Note, here we have to specify a new data table for the plot to use
base_plot + 
    geom_ribbon(data = expt1_deg2_means_summary,aes(ymin = lower.CL,ymax = upper.CL),alpha = 0.2) + 
    geom_line(data = expt1_deg2_means_summary,aes(y = emmean),size=1.5)

```

We will consider two alternate designs for this experiment in the problems below

# Experiment 2

In an attempt to better characterize the Nitrogen response trend, 
the researcher modifies the experimental design to be the following:


- 10 nitrogen regimes: 0, 45, 90, 120, 150, 180, 210, 240, 270, 300 lbs / acre at emergence
- 5 rep (plot) / Nitrogen level
- Response: total yield

The data for this experiment is here:
```{r}
experiment2 = read.csv('Experiment2.csv',stringsAsFactors=F)
str(experiment2)
```


## 1.1 Prepare the design table for Experiment 2

**Note**: I recommend creating the table in Excel, copy/paste it into: https://www.tablesgenerator.com/markdown_tables under `File -> Paste table data...`, then copy the markdown and paste here:


| Structure | Variable  | Type        | # levels | EU        |
|-----------|-----------|-------------|----------|-----------|
| Treatment | Nitrogen  | Numeric     | 10       | NitrogenF |
| Design    | NitrogenF | Categorical | 10       |           |
|           | Plot      | Categorical | 50       |           |
| Response  | Yield     | Numeric     | 50       |           |

> [5 points] Note here 1) # levels of all terms. 2) Nitrogen is Numeric (or Quantitative), and NitrogenF (or similar name) is Categorical (or Factor)
3) EU for Nitrogen is NitrogenF


## 1.2 Does this experiment require more work to carry out than Experiment 1? Why or why not?

> It shouldn't require any more work. Ideally the Nitrogen formulation is created individually for every plot, so it has to be done 50 times whether there are 5 reps/level of 10. However, since the experimental unit is the level of nitrogen itself, the design wouldn't change if each concentration was set only once and re-used on all replicate plots at that level. In this case we would have to declare NitrogenF as the EU even if we wanted to compare levels categorically, and so each level would be un-replicated. But here, we'd still have 10 EUs. If we did it this way, we would have to make 10 formulations instead of 5, but the work of growing and harvesting the potatoes would be the same.

## 1.3 Fill in the missing values (denoted by ??) of the following ANOVA table
This is calculated for a trend on Nitrogen modeled with a polynomial. What was the degree of the polynomial?

| Source   | Df | Sum_of_sq | Mean_sq | F-value | p-value |
|----------|----|-----------|---------|---------|---------|
| Nitrogen | 3  | 21980.4   | 7326.8  | 4.4398  | 0.05735 |
| Error    | 6  | 9901.5    | 1650.25 |         |         |

> The degre of the polynomial here was 3 (ie I(N)+I(N^2)+I(N^3))

> The DfE is #EU - DfT - 1
> The Mean squares are calculated as SSX/DfX where X=Treatment or Error

> [5 points] -1 for each missed calculation

## 1.4 Choose a better polynomial trend model (if possible) based on these data and plot the fit + confidence ribbon

Here is some more background on choosing the best polynomial trend model:

The goal is to choose a model with enough, but not too many parameters (ie just the right degree). If the degree is too small, the model isn't flexible enough to fit the real trend of the data, **ie it will give a biased prediction for some levels of Nitroge**. As the degree increases, a polynomial function can approximate almost any curve shape. But, higher degree means large DfT, potentially a wigglier (and less realistic) trend-line, and also the potential to over-fit the data. Over-fitting is similar to under-estimating confidence intervals and leads us to predict small-scale patterns (like increases-then decreases-then increases in Yield as we increase N) when they are not there.

I recommend you start with the following guidelines:

1. Choose a maximum degree for your model to start with:
  - The maximum degree MUST be less than the number of EU. I wouldn't go higher than #EU-3, and wouldn't even go that high unless really needed.
  - Unless your treatment covers a very large range and you have a lot of data, I wouldn't try degrees > 5 or 6. Degree-5 polynomials already do a pretty good job of approximating most realistic response functions
2. Fit this model by spelling out each exponent of your treatment (`~I(Variable)+I(Variable^2)+I(Variable^3)+...`) and call it a "full trend". This will be your reference for a model of maximum complexity.
3. Use the `anova` function to decide if the `full_trend` model is necessary.
  - If you use `lmer` to fit your model, be sure to set the argument `ddf='Kenwood-Roger'` (`ddf = 'K'` for short), and the argument `type='I'`. 
  - The ANOVA will break up the Treatment model into a separate row for each exponent. Look specifically at the row corresponding to the highest power (it should be the last row). This row of the table constructs an F-statistic to test the null hypothesis that this highest power of your model is NOT IMPORTANT (i.e. $\beta_k = 0$ where $k$ is the highest power in your trend model). If these highest power is not important, you'd be better off dropping it. Make a decision based on the F-statistic (interpretation: larger F-statistics mean the term is more important. For example F_hat = 3 means that this exponent is roughly 3x as important as the deviations of level-means from your curve). But remember that F-statistics can be large by chance so the hypothesis test (p-value) can guide you whether this large an F could be by chance.
4. If the highest exponent is NOT SUFFICIENTLY important, then drop this term from your model, call it a `reduced_trend`, and re-fit. You'll get a benefit from more degrees of freedom for error without compromising much in the trend itself.
5. Repeat 3-4 until the highest remaining power of your treatment is deemed **important** by the ANOVA.
6. Perform model diagnostics on this final model using the `pls205_diagnostics()` function.

The most difficult step here is Step 3. There's an art to this, as much as a science. The easiest strategy is to simply choose an $\alpha$ threshold, and apply this to the p-value of the highest power term each time. I recommend that you choose a fairly liberal one like 0.1. If you end up with too-complex a model, your confidence intervals will be bigger. You do not need to correct for multiple testing in this context (at least for this course, your PI may have a different opinion but it is a judgment call).

Present your full model selection in code blocks below. For maximum transparency, I recommend using a different code block for each time through steps 3-5 and ending with the ANOVA for that model, and then making a comment after the code block describing why you chose to drop the highest-remaining term or not. You can receive full credit for a well-reasoned choice even if it doesn't match the answer key exactly. End with your model diagnostics and a plot showing the fitted trend-line and confidence intervals as done above (reuse that code, but be sure to change the inputs!).

```{r}
# code for full model
experiment2$NitrogenF = as.factor(experiment2$Nitrogen)
full_trend = lmer(Yield ~ I(Nitrogen)+I(Nitrogen^2)+I(Nitrogen^3)+I(Nitrogen^4)+I(Nitrogen^5)+(1|NitrogenF),data = experiment2)
anova(full_trend,ddf = 'K',type='I')
```

> I chose a degree-5 model to start. This already leaves only 4 DfE which is not very many to fit 5 model parameters. 
> From the anova table, the N^5 term looks very unimportant. The F-value < 1 meaning it explains very little amplitude in the curve, and the p-value is > 0.1. So I will drop this term and re-fit

```{r}
#code for first reduced model
trend4 = lmer(Yield ~ I(Nitrogen)+I(Nitrogen^2)+I(Nitrogen^3)+I(Nitrogen^4)+(1|NitrogenF),data = experiment2)
anova(trend4,ddf = 'K',type='I')
```

> I will keep this as my final model. The N^4 term does look important (maybe not as important at the N^2 term, but it does explain 17x as much of the variation as the error). The p-value for this term suggests that the effect of N^4 is significant (non-zero), so it appears to be important to keep.
> We never drop lower-order terms than the first "significant" exponent, so we'll stop here.

```{r}
# code for diagnostics
pls205_diagnostics(trend4,EU = 'NitrogenF')
```

> The diagnostics lookOK. The qqplot loks pretty good, expect maybe point 7. But there are so few points, it's really hard to tell. The S/L plot may suggest some problems with lower deviations at lower fitted values. It might be worth trying transformation (maybe a power transformation?), but there are so few EU that it's a bit hard to be sure. It's probably safe to continue with this model

```{r}
# calculate the trend at a range of values of Nitrogen:
Nitrogen_range = seq(0,320,by=1)
expt2_trend4_means = emmeans(trend4,spec = 'Nitrogen',at=list(Nitrogen = Nitrogen_range))
expt2_trend4_means_summary = as.data.frame(summary(expt2_trend4_means,infer = c(T,F),level = 0.95,ddf='K'))
# code for plotting
base_plot = ggplot(experiment2,aes(x=Nitrogen)) + 
    geom_boxplot(aes(y=Yield,group = Nitrogen),color = 'blue',position = position_identity()) + 
    ggtitle('Alpine Russet') + expand_limits(y=0)

# Then, add a line for the trend and a ribbon for the confidence intervals.
# Note, here we have to specify a new data table for the plot to use
base_plot + 
    geom_ribbon(data = expt2_trend4_means_summary,aes(ymin = lower.CL,ymax = upper.CL),alpha = 0.2) + 
    geom_line(data = expt2_trend4_means_summary,aes(y = emmean),size=1.5)
```

> The trend fits the observed data pretty well. 

> [10 points] 
     - 5 points for full model including a polynomial for N and a random effect for (1|NitrogenF), and using lmer. It is OK if you started with anywhere from 4-7 as the starting degree
     - 2 points for explaining whether the last term should be kept of not from the full model. (-1) for dropping multiple terms at once
     - 1 point for showing and interpreting the diagnostics
     - 1 point for getting the visualization code to work


## 1.5 Compare (visually) the trend estimated from Experiment2 to the trend estimated from Experiment1
Is the general shape of the trend similar? Is the width of the confidence interval band the same?
Which experiment provided more information about the response of Yield to Nitrogen? 
How can you tell?

> The center-part of the two trend-lines are fairly similar, peaking ~150lbs or so. But the trend from experiment 2 predicts a steeper decline going away from this middle value, and also seems to predict a bit of an up-tick as you get close to zero or towards very large values. 
> The width of the confidence interval bands are much narrower based on the data from experiment 2. This is a result of fitting the data better (less error), and because there were more EUs with a narrower spacing across the range.
> Experiment2 clearly provides more information because of the smaller confidence intervals.

> [5 points] 2 points for comparing shape, 2 for comparing CIs, 1 for explanations

# Experiment 3

If Experiment 2 was better than Experiment 1, why not go further and use a different level of Nitrogen in every plot? Experiment 3 has the following design:

- 50 nitrogen regimes distributed equally from 0 to 300 lbs / acre at emergence
- 1 rep (plot) / Nitrogen level
- Response: total yield


The data for this experiment is here:
```{r}
experiment3 = read.csv('Experiment3.csv',stringsAsFactors=F)
str(experiment3)
```

## 2.1 Prepare the design table for Experiment3
 
| Structure | Variable  | Type        | # levels | EU        |
|-----------|-----------|-------------|----------|-----------|
| Treatment | Nitrogen  | Numeric     | 50       | NitrogenF |
| Design    | NitrogenF | Categorical | 50       |           |
|           | Plot      | Categorical | 50       |           |
| Response  | Yield     | Numeric     | 50       |           |

> Note here 1) # levels of all terms. 2) Nitrogen is Numeric (or Quantitative), and NitrogenF (or similar name) is Categorical (or Factor)
3) EU for Nitrogen is NitrogenF
> It is OK if Plot is excluded because it is aliased with NitrogenF.

> [5 points] 

## 2.2 Following the same strategy as above, select a polynomial trend model for Yield~f(Nitrogen) based on the data from Experiment3
Be careful about specifying this model using all rules about turning a design table into a model in R!

```{r}
experiment3$NitrogenF = as.factor(experiment3$Nitrogen)
full_trend_expt3 = lm(Yield ~ 1 + I(Nitrogen)+I(Nitrogen^2)+I(Nitrogen^3)+I(Nitrogen^4)+I(Nitrogen^5),data = experiment3)
anova(full_trend_expt3)
```
> Note: we cannot include the EU in our model because it has the same # levels as the response.
> I started with a degree-5 again because going higher is rarely needed or justified biologically
> Based on this, it appears we are safe dropping N^5 again

```{r}
trend4_expt3 = lm(Yield ~ 1 + I(Nitrogen)+I(Nitrogen^2)+I(Nitrogen^3)+I(Nitrogen^4),data = experiment3)
anova(trend4_expt3)
```
> Once again, it appears that the N^4 term is important so we will stop

```{r}
pls205_diagnostics(trend4_expt3)
```
> Diagnostics look good. There's a bit of a hump in the middle of the S/L plot which we might be a bit concerned about given the moderate amount of data, but it really doesn't look very severe to me.

```{r}
# calculate the trend at a range of values of Nitrogen:
Nitrogen_range = seq(0,320,by=1)
expt3_trend4_means = emmeans(trend4_expt3,spec = 'Nitrogen',at=list(Nitrogen = Nitrogen_range))
expt3_trend4_means_summary = as.data.frame(summary(expt3_trend4_means,infer = c(T,F),level = 0.95))
# code for plotting
base_plot = ggplot(experiment3,aes(x=Nitrogen)) + 
    geom_boxplot(aes(y=Yield,group = Nitrogen),color = 'blue',position = position_identity()) + 
    ggtitle('Alpine Russet') + expand_limits(y=0)

# Then, add a line for the trend and a ribbon for the confidence intervals.
# Note, here we have to specify a new data table for the plot to use
base_plot + 
    geom_ribbon(data = expt3_trend4_means_summary,aes(ymin = lower.CL,ymax = upper.CL),alpha = 0.2) + 
    geom_line(data = expt3_trend4_means_summary,aes(y = emmean),size=1.5)
```


> [10 points] 
     - 5 points for full model including a polynomial for N but NO random effect (because NitrogenF has the same number of levels as the response), and using lmer. It is OK if you started with anywhere from 4-7 as the starting degree
     - 2 points for explaining whether the last term should be kept of not from the full model. (-1) for dropping multiple terms at once
     - 1 point for showing and interpreting the diagnostics
     - 1 point for getting the visualization code to work
     - Note: it is OK if you ended up with a different degree than me.


## 2.3 Compare (visually) the trend estimates from Experiment 2 and Experiment 3.
Is there any reason that you can think of of favoring Experiment2 over Experiment 3 in this context?

> The curves from experiment 2 and experiment 3 are very similar. But as expected, the confidence intervals are even narrower from experiment 3 suggesting that this experiment is more powerful than experiment2 for modeling the trend. Also note that this one finds less evidence for a strong up-tick for very small or very large values, thought it's still there a bit. This is a consequence of the degree-4 trend. Had we used a degree-5, it would have ticked up on one side and down on the other. This shows that it is somewhat dangerous to make too much out of the fits towards the boundaries of the data; the model fits in the central part of the data are generally much more robust and trustworthy.

> One reason to favor Experiment2 would be because it also allows you to compare specific N levels. Experiment 3 is un-replicated at any specific N level, so while you can use the trend itself, you can't directly compare 150lbs to 0lbs because each level is un-replicated. You can make this comparison indirectly using the model and emmeans like this:

```{r}
expt3_trend4_means = emmeans(trend4_expt3,specs = 'Nitrogen',at=list(Nitrogen = c(150,0)))
pairs(expt3_trend4_means)
```

> which gives a similar estimate to what we would get if we analyzed Experiment 2 as a categorical variable

```{r}
expt2_categorical = lm(Yield ~ NitrogenF,experiment2)
exp2_means_cat = emmeans(expt2_categorical,specs = 'NitrogenF',at=list(NitrogenF = c('150','0')))
pairs(exp2_means_cat)
```

> using the trend is somewhat more complicated and subject to how much we really want to trust the trend.

> [5 points] 4 points for the comparison, 1 for tackling the last qustion. The analysis I did above is not required.
