---
title: "Lab 3. ANOVA and linear models"
author: Daniel Runcie.
output: 
 html_notebook:
    toc: true
    toc_float: true
---

``` {r}
library(ggplot2)
library(emmeans)
library(car)
library(lme4)
library(lmerTest)
library(multcomp)
library(multcompView)
```

## Experiment with more than two treatments

In the following experiment, the nitrogen fixation capacities of six different strains of rhizobia on clover are compared. 
The experiment is arranged as a completely randomized design (CRD) with 6 treatments (i.e. 5 new strains plus a control) 
and five independent replications (e.g. plots) per treatment.

**Design**: Completely randomized design

| Structure | Variable | Type        | # levels | Experimental Unit |
|-----------|----------|-------------|----------|-------------------|
| Treatment | Strain   | Categorical | 6        | Plot              |
| Design    | Plot     | Categorical | 30       |                   |
| Design    | Sample   | Categorical | 120      |                   |
| Response  | NLevel   | Numeric     | 120      |                   |


> Thought Question:
>
> What types of comparisons among treatments would be most interesting? 

- We could compare each new strain to the control. 
- Or we could compare all pairs of strains.
- Or we could test the hypothesis that all strains are equivalent.

First, Load the data and check that it was loaded correctly:
```{r}
clover.dat <- read.csv('~/Desktop/R_Projects/PLS205_Winter2021/PLS205_Winter2021/data/Clover_data.csv')
str(clover.dat)
summary(clover.dat)
```

`Culture` is a Factor with 6 levels, and `NLevel` is numeric data.
`Plot` should be a factor with 30 levels, but is an `int` with 5 values. It's not important
for this analysis, but in general we should fix this and **Nest** Plot within culture.
Nesting means giving each Plot a unique name within each culture. We can nest using the `interaction()` 
function.

```{r}
clover.dat$Plot <- as.factor(clover.dat$Plot)
clover.dat$Strain <- as.factor(clover.dat$Strain)
clover.dat$Sample <- as.factor(clover.dat$Sample)
str(clover.dat)

clover.dat$Plot <- interaction(clover.dat$Strain,clover.dat$Plot)
clover.dat$Plot

```

## Analyzing the data

As we showed in Lab 2, data analysis in R consists of three steps:

1. Visualize the data
2. Fit a model to the data
3. Run model diagnostics to check model assumptions
4. Use the model fit to calculate summary statistics and draw conclusions

### Visualize the data
Let's start by plotting the data to visually inspect
```{r}
library(ggplot2)
ggplot(clover.dat,aes(x=Strain,y=NLevel)) + geom_boxplot() + geom_jitter()
```

It looks like we do have large differences among the treatments. Let's see if the statistics
backs this up.


### Fit the model:

Since the experimental design of this experiment is also a CRD with only a single class of treatments,
the form of the model is the same as for the Barley data. 

- The **explanatory** component is the Culture (ie the species of bacteria/ the treatment)
- The **error** component is caused by variation among pots, and measurement error. This is all just variation
due to the sampling of experimental units in a CRD, so we don't write it in our model

Therefore, the model statement looks the same as before:

```{r}
clover.mod <- lm(NLevel ~ Strain, data = clover.dat)
clover.mod
```

You'll notice that we now have more **Coefficients** in the brief model summary.

### Run diagnostic plots
```{r}
op = par(mfrow=c(1,2))
plot(clover.mod,which = c(2,3))
```

With more complex designs, The **Scale-Location** plot tends to be more useful than the **Residuals vs Factor Levels**
plot we used in Lab 2. The **Scale-Location** plot shows the absolute value of standardized residuals against the fitted values.
In this case, fitted values are the treatment means (ie the explanatory component of the model). 
You should check for a trend such as the magnitude of residuals increasing for larger fitted values (a common problem). The
red line tries to help visualize this trend.

We do see some problems here. The residuals are a bit skewed (too many small values), 
and the cultures seem to vary some in their variances, 
with the cultures with higher Nitrogen levels having slightly higher residuals variances. 
However, neither looks very severe.

We might want to stop here, or consider data transformations (a later lab). But we'll continue on for demonstration.

### Generate model summaries and tests
Most of these analyses can be run using the `emmeans` function:
```{r}
library(emmeans)
```


#### Treatment means
The first analysis might be to simply estimate the mean Nitrogen Level for each of the 6 Cultures.
```{r}
means_clover = emmeans(clover.mod,~Strain)  # This preps an emmeans analysis, grouping by Culture
summary(means_clover,level = 0.95,infer = c(T,F))  # note setting level (1-alpha) for the confidence intervals
# infer = c(T,F) gives confidence intervals
# infer = c(F,T) gives p-values intervals. Here these are testing against the null hypothesis that mu_i = 0
# infer = c(T,T) gives both.
```

#### Treatment differences
With 6 levels of the treatment, there are several ways of describing differences.

1. We could compare each of the Cultures to a single control treatment
2. We could compare all pairs of Cultures

#### Confidence intervals and hypothesis tests for differences between new cultures and control
We tell `emmeans` to measure the effects of each treatment against the control with the command `trt.vs.ctrl~Culture`, 
and then specify the control treatment with `ref = 6` because `Comp` is the 6th treatment as listed above. 
You can see how R orders treatments like this:
```{r}
levels(clover.dat$Culture)
```

"Comp" is the 6th level of the variable `Culture` in the data.frame `clover.dat`

```{r}
differences_vs_control = emmeans(clover.mod,trt.vs.ctrl~Strain,ref = 6)  # ref = 6, because the control is the 6th Culture
summary(differences_vs_control,level = 0.95,infer = c(T,T))$contrast
# by adding $contrast after the summary function, we get only the contrast table, not the treatment means table
```

Note: The summary states that the **dunnetts** method was used to adjust the level of the confidence intervals.
This is the correct adjustment for comparisons of treatments against a single control. 

This adjustment works by calculating a new **penalty** statistic to replate $t_{\alpha/2,df}$ in
the formula for a confidence interval.

The **p.value** for each test is also adjusted so that it is a **simultaneous** p-value for all the tests.
It represents the probability that the **biggest** difference among any of the 5 `3DOk` species against
the control would be at least as large as the biggest observed difference **if all 5** were the same
as the control.


#### Confidence intervals for pairwise differences among all cultures
We tell `emmeans` to form all pairwise comparisons with `pairwise ~ Culture`
```{r}
all_differences = emmeans(clover.mod,pairwise~Strain)
summary(all_differences,level = 0.95,infer = c(T,T))$contrast
```

#### Exercise: Form the treatments into groups
Making sense of all these pairwise comparisons (also called **contrasts**) is challenging.

One common strategy is to form the treatments into groups, where all treatments that cannot
be significantly distinguished are in the same group, and all treatments that can be distinguished
(at a specific alpha) are in different groups. Commonly groups are assigned letters ("a","b","c", etc)
or numbers ("1","2","3").

By "distinguished" here, we mean that we can reject the **null hypothesis** that $\mu_i = \mu_j$,
and that the chance of incorrectly stating that **any pair** of treatments is different when they are
actually all the same is $\alpha$*100.

The strategy for forming these groups is the following:

1. Find the **minimum significant difference** that can be distinguished. 
Remember, a difference is **significant** if its confidence interval does not cross zero.
The confidence interval is $\hat{\delta} \pm t_c * SED$ (mouse-over to see the equation rendered).
So, two treatments are different if $|\hat{\delta}| > t_c * SED$. 
  - find SED (*hint: look at the pairwise difference table above*)
```{r}
SED = 2.171513  # replace by the correct number.
```
  - Find t_c. This will come from the Tukey distribution instead of the t-distribution. The Quantile 
  function of this distribution is `qtukey`. Alternatively, you could use the confidence intervals
  above to calculate it!
```{r}
# Note: using qtukey is a bit different than qt. 
#   You also have to enter the number of groups (nmeans)
#   The value that you give out needs to be divided by sqrt(2)
#   The tukey distribution is "folded", so it counts negative and positive differences the same.
#   This means we use it like a one-sided distribution, and use p=alpha (instead of p=alpha/2) 
c_q = qtukey(p = 0.05,nmeans = 6,df = 24,lower.tail = F) / sqrt(2)
c_q
# instead, from the confidence interval of the first contrast above: (22.27416704-8.845833)/(2*SED)
```
  - The minimim significant difference is then `c_q * SED`
```{r}
MSD = c_q * SED
MSD
```
  
2. Order the treatments from smallest to largest by their mean (`emmean`)
| # | Treatment | Mean | Letters |
|---|-----------|------|---------|
| 1 |           |      |         |
| 2 |           |      |         |
| 3 |           |      |         |
| 4 |           |      |         |
| 5 |           |      |         |
| 6 |           |      |         |

3. Start with the smallest treatment. Assign this an **a**. Then moving up the list of treatments,
label all treatments with an **a** if they are within `MSD` of the first treatment.
4. When you get to a treatment that can be distinguished from the first, label this treatment **b**.
5. Now, find the first treatment (from the top) that is NOT distinguishable from the first **b**. Add a **b** to this treatment.
Note: treatments can receive multiple letters!
6. Now, repeat step 3 starting with the smallest **b** treatment.
7. Repeat 4-6 for **c**, **d**, etc, until you label the last treatment.

Do this on paper or in the table above. Work with those around you.


#### Compact letter display with emmeans
This is a bit of work. Fortunately, the `emmeans` package has a function `cld` that will do this for you!
```{r}
cld(all_differences$emmeans,Letters = letters)
```

#### ANOVA
If you can distinguish two treatments from each other, this means that you can conclude that 
the treatment has an effect. However, it is sometimes cumbersome to have to search through a long 
list of pairwise contrasts to check if any of them are *significant*. Sometimes, it is
convenient to get a simple statistic that states how likely it would be to see your data if all
treatments were ineffective (ie all means were the same). An ANalysis Of VAriance can provide this.

To run an ANOVA for a simple model, do this:
```{r}
anova(clover.mod)
```

The **F value** is similar to the **t-statistic** in a t-test. It is a standardized measure of 
the among-treatment differences relative to the within-treatment variation. F close to 1 means 
the treatments are similar, while large F values signify large differences among treatments. The
corresponding p-value measures the probability of this large (or larger) differences in treatment
means if all treatments were identical. In this case, we have strong evidence that 
**at least one** treatment differs from the others. Which? We'd have to go back to the pairwise
differences test to know.

---

## Data entry problems

It is **very** important to carefully inspect how R loaded your data. 

I've taken the exact same clover data and saved it in a new file (`Clover_data_Renamed.csv`), but named the Cultures 1:6. 

Load this data and repeate the abova ANOVA. Do you get the same answer? What is wrong?
```{r}
# load data and view
clover.dat.renamed <- read.csv("~/Desktop/R_Projects/PLS205_Winter2021/PLS205_Winter2021/data/Clover_data_renamed.csv")
str(clover.dat.renamed)
```
> **Note** Culture is an `int`, not a `Factor`! 

Proceed anyway and fit the model try to get treatment means
```{r}
clover.mod.v2 <- lm(NLevel ~ Strain, data = clover.dat.renamed)

clover_means_2 = emmeans(clover.mod.v2,~Strain)
summary(clover_means_2,level = 0.95,infer = c(T,F))
```

This is basically meaningless (an estimate of the mean weed biomass for Clover 3.5).

You can fix this by running this right after loading the data:
```{r eval = F}
clover.dat.renamed$Strain = as.factor(clover.dat.renamed$Strain)
clover.mod.v3 <- lm(NLevel ~ Strain, data = clover.dat.renamed)


clover_means_3 = emmeans(clover.mod.v3,~Strain)
summary(clover_means_3,level = 0.95,infer = c(T,F))
```

-----------------

## Power analysis of an ANOVA
A power analysis of an experiment with >2 treatments works very similarly to a power analysis of a t-test. `R`'s function is: `power.anova.test`. 

- `groups` the number of groups (or treatments)
- `n` the number of samples **per group**. This only works if the two groups have the same number of samples
- `between.var` the variance of the group means
- `within.var` the **true** variance within each group (assumes same across groups)
- `sig.level` the desired Type I error probability
- `power` the desired power (1-TypeII error probability)

The main differences from `power.t.test` are that we now have to specify the number of groups, and that `delta` and `sd` have been replaced by `between.var` and `within.var`. `within.var` is simply $\sigma^2$. `between.var` is the variance of **group means**. This is tricky. How can you specify what **variance** is interesting?

The easiest way to do this is to specify the minimum difference between any two treatments that would be interesting, and test the situation that the first two treatments differ by that amount, and that the other treatment means are all equal to the global mean. This would give the **smallest interesting** treatment effect, and so the one that is hardest to detect.

Let's say we care about a difference of 50 units with our five `Nitrogen` treatments:
```{r}
overall_mean = 400  # this is arbitrary
delta = 50

# create a vector with 6 hypothetical means, with the first two differing by 5 and the rest all equal to the global mean
hypothetical_means <- overall_mean + c(delta/2,-delta/2,0,0,0,0)

# calculate the between.var of this minimally interesting set of means
between.var <- var(hypothetical_means)
between.var
```

Now, lets use this to run a power analysis. We have to specify the following:

- sample size = 10 reps / treatment
- within-treatment variance = 2500 (we assume this is known)
- alpha = 0.05
```{r}
n = 10
s2 = 2500
alpha = 0.05
power.anova.test(groups = 5,
                 n = n,
                 between.var = between.var,
                 within.var = s2,
                 sig.level = alpha,
                 power = NULL)
```

> **Interpretation:** With this experiment, we would have roughly a 30% chance of getting a significant result if two means differed by 50 and the rest were all equal to the global mean.

### Activites
Explore the effects of several modifications to the power analysis:

1. Test the power if the two most extreme treatments differ by 50, but the other 3 means vary between these means: (ex. c(380,500,420))
2. Test the power if the number of groups go up or down, but the `between.var` and `within.var` stay the same
3. Test the power if `n` or `within.var` change.
4. Compare the power to that of a t-test (just two groups that differ by 50), but with the same `n`, `within.var`, and `alpha`.

Do these follow your expectations?



